---
title: "Reproducible Research. Peer Assessment 1"
author: "Roman Savchenko"
date: "2022-12-24"
output: html_document
---


## Change the locale for the R process.
```{r setlocale, echo = TRUE}
Sys.getlocale()
Sys.setlocale("LC_ALL","English")
Sys.getlocale()
```


## Loading and preprocessing the data

The reading of the data from the activity.csv file to the R data frame.
```{r loaddata, echo = TRUE}
activity <- read.csv(file = 'activity.csv')
head(activity)
str(activity)
```

The creation of an additional variable to count cases when activity$steps=NA.
```{r addnavar, echo = TRUE}
activity$stepsNA <- ifelse(is.na(activity$steps), 1, 0)
head(activity)
str(activity)
unique(activity$stepsNA)
```

The creation of an additional variable for 2 digit value of the month.
```{r addmonth, echo = TRUE}
activity$month <- substr(activity$date, 6, 7)
head(activity)
str(activity)
unique(activity$month)
```

The creation of an additional variable for the numeric date variable.
```{r addndate, echo = TRUE}
activity$dateN <- as.Date(activity$date, format = "%Y-%m-%d")
head(activity)
str(activity)
```

The creation of an additional variable for the day of the week.
```{r addweekday, echo = TRUE}
activity$weekDay <- weekdays(activity$dateN, abbreviate = TRUE)
head(activity)
str(activity)
unique(activity$weekDay)
```

The creation of an additional factor variable for the day of the week.
```{r addweekday_fact, echo = TRUE}
activity$weekDayFact <- as.factor(ifelse(activity$weekDay %in% c("Sat", "Sun"),
                                         "weekend", "weekday"))
head(activity)
str(activity)
unique(activity$weekDayFact)
```

The including of the R Stats package for the calculations.
```{r incl_stats, echo = TRUE}
library(stats)
```

The including of the R Graphics package to produce histogram.
```{r incl_graph, echo = TRUE}
library(graphics)
```

The including of the R package to produce time series plot.
```{r incl_ggplot2, echo = TRUE}
library(ggplot2)
```


## What is mean total number of steps taken per day?

The calculation of the total number of steps taken per day.
```{r steps_day, echo = TRUE}
stepsPerDay <- aggregate(x = activity$steps, by = list(activity$date), 
                         FUN = sum, na.rm = TRUE)
    colnames(stepsPerDay)[colnames(stepsPerDay) == 'Group.1'] <- 'date'
    colnames(stepsPerDay)[colnames(stepsPerDay) == 'x'] <- 'StepsTot'
print(stepsPerDay)
```

The creation of a histogram of the total number of steps taken each day.
```{r hist_totsteps, echo = TRUE}
hist(stepsPerDay$StepsTot,
     main = "Histogram of The Total Number of Steps Taken Each Day",
     breaks = 25,
     xlim = c(0, 25000),
     ylim = c(0, 12),
     xlab = "Total Number of Steps",
     labels = TRUE)
```

The calculation of the mean of the total number of steps taken per day.
```{r steps_mean, echo = TRUE}
MeanSteps <- mean(stepsPerDay$StepsTot, na.rm = TRUE)
MeanSteps
MeanStepsRnd <- round(MeanSteps, digits = 2)
MeanStepsRnd
```

The calculation of the median of the total number of steps taken per day.
```{r steps_median, echo = TRUE}
MedianSteps <- median(stepsPerDay$StepsTot, na.rm = TRUE)
MedianSteps
MedianStepsRnd <- round(MedianSteps, digits = 2)
MedianStepsRnd
```


## What is the average daily activity pattern?
The making of a time series plot of the 5-minute interval and the average number of steps taken.
```{r timeseries_plot, echo = TRUE}
stepsPerInterval <- aggregate(x = activity$steps, by = list(activity$interval), 
                              FUN = mean, na.rm = TRUE)
    colnames(stepsPerInterval)[colnames(stepsPerInterval) == 'Group.1'] <- 'interval'
    colnames(stepsPerInterval)[colnames(stepsPerInterval) == 'x'] <- 'StepsAvg'
plot <- ggplot(stepsPerInterval, aes(x = interval, y = StepsAvg)) +
        geom_line() +
        labs(x = "5-minute Interval", y = "Avereage of the steps", 
             title="Avereage of number of steps taken per 5-minue intervals")
plot
```

The selection of the 5-minute interval with the maximum number of steps
```{r maxstep_interval, echo = TRUE}
MaxStepAvg <- max(stepsPerInterval$StepsAvg)
MaxInt <- subset(stepsPerInterval, StepsAvg == MaxStepAvg,
                 select = c(interval, StepsAvg))
MaxInt
```


## Imputing missing values

The calculation of the total number of missing values in the data frame.
```{r total_missing, echo = TRUE}
sum(activity$stepsNA)
```

The missing values will be imputed with the average value for the same time interval
```{r impute_missing, echo = TRUE}
PreImp <- merge(activity, stepsPerInterval, by = "interval")
StepsImp <- PreImp
   StepsImp$steps <- ifelse(is.na(StepsImp$steps), StepsImp$StepsAvg, StepsImp$steps)
head(StepsImp)
str(StepsImp)
```

The creation of a histogram of the total number of imputed steps.
```{r hist_impsteps, echo = TRUE}
StepsImpDay <- aggregate(x = StepsImp$steps, by = list(StepsImp$date), 
                         FUN = sum, na.rm = TRUE)
    colnames(StepsImpDay)[colnames(StepsImpDay) == 'Group.1'] <- 'date'
    colnames(StepsImpDay)[colnames(StepsImpDay) == 'x'] <- 'StepsImpTot'
print(StepsImpDay)
hist(StepsImpDay$StepsImpTot,
     main = "Histogram of The Total Number of Imputed Steps Taken Each Day",
     breaks = 25,
     xlim = c(0, 25000),
     ylim = c(0, 25),
     xlab = "Total Number of Steps",
     labels = TRUE)
```

The calculation of the mean of the total number of steps taken per day.
```{r stepsimp_mean, echo = TRUE}
MeanImpSteps <- mean(StepsImpDay$StepsImpTot, na.rm = TRUE)
MeanImpSteps
MeanImpStepsRnd <- round(MeanImpSteps, digits = 2)
MeanImpStepsRnd
```

The calculation of the median of the total number of steps taken per day.
```{r stepsimp_median, echo = TRUE}
MedianImpSteps <- median(StepsImpDay$StepsImpTot, na.rm = TRUE)
MedianImpSteps
MedImpStepsRnd <- round(MedianImpSteps, digits = 2)
MedImpStepsRnd
```

The mean value is changed from `r MeanStepsRnd` to `r MeanImpStepsRnd` and the median is changed from `r MedianStepsRnd` to `r MedImpStepsRnd`.


## Are there differences in activity patterns between weekdays and weekends?

Splitting data for weekdays and weekends in different data frames.
```{r split_weekdays, echo = TRUE}
WeekDays <- subset(activity, weekDayFact == "weekday")
WeekEnds <- subset(activity, weekDayFact == "weekend")
```

Calculation of the average number of steps for each new data frame.
```{r }
WeekDaysInt <- aggregate(x = WeekDays$steps, by = list(WeekDays$interval), 
                         FUN = mean, na.rm = TRUE)
    colnames(WeekDaysInt)[colnames(WeekDaysInt) == 'Group.1'] <- 'interval'
    colnames(WeekDaysInt)[colnames(WeekDaysInt) == 'x'] <- 'StepsAvg'
plot <- ggplot(WeekDaysInt, aes(x = interval, y = StepsAvg)) +
        geom_line() +
        labs(x = "5-minute Interval", y = "Avereage of the steps", 
             title="Avereage of number of steps across weekdays")
plot

WeekEndsInt <- aggregate(x = WeekEnds$steps, by = list(WeekEnds$interval), 
                         FUN = mean, na.rm = TRUE)
    colnames(WeekEndsInt)[colnames(WeekEndsInt) == 'Group.1'] <- 'interval'
    colnames(WeekEndsInt)[colnames(WeekEndsInt) == 'x'] <- 'StepsAvg'
plot <- ggplot(WeekEndsInt, aes(x = interval, y = StepsAvg)) +
        geom_line() +
        labs(x = "5-minute Interval", y = "Avereage of the steps", 
             title="Avereage of number of steps across weekends")
plot
```
